\thetitle

### 2009-02-24 paper/paper.tex
Final tweaks (honestly).
Ignore-this: 8f6fbcf924041e9f7123a040b692ccf

darcs-hash:20090224144219-bab43-41dddddf608c66f18651c9380b6288affda59c6e.gz

The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
    was 
%%% could be 
fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.

The left inequality is the challenging one. To bring it closer to the index summation, we first write |l| on the left side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).

@@

Again, the proof of (\ref{monoid_expec_IJ_lem}) 
is by induction on |qs|. But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
\begin{code}
  forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
    monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
\end{code}
In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
\begin{code}
monoid_expec (count (i, j)) (
    pi <- pick [0 ... n-1];
    let pivot := nth li pi in
    part <- partition_U pivot (remove li pi);
    lower <- qs_U (part Lt);
    upper <- qs_U (part Gt);
    ret (lower ++ pivot :: part Eq ++ upper)
  ) <= ecc i j.
\end{code}
Since |cmp_U| is deterministic, |partition_U| is as well. Furthermore, since
we know exactly what monadic effects |partition_U| has, we can split those
effects off and revert to simple effect-free |filter| passes. Finally, we rewrite using the following |monoid_expec| identity:
\begin{code}
  monoid_expec f (pick l >>= m) = avg (map (monoid_expec f fcmp m) l).
\end{code}
This way, the goal ends up in a form using less monadic indirection:
\begin{code}
avg (map (monoid_expec (count (i, j)) fcmp (fun pi =>
    let pivot := nth li pi in
    let rest := remove li pi in
    let flt := fun c => filter ((= c) fcmp lookup_cmp pivot) rest in
    ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
      lower <- qs_U (flt Lt);
      upper <- qs_U (flt Gt);
      ret (lower ++ pivot :: flt Eq ++ upper)
    ))) [0...n-1]) <= ecc i j.
\end{code}
Here, |map_fst| applies a function to a pair's first component.

We now distinguish between five different cases that can occur for the nondeterministically picked |pivot| (which, because we are in the |U| monad, is an index). It can either be less than |i|, equal to |i|, between |i| and |j|, equal to |j|, or greater than |j|. Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls). To represent this split, we first rewrite the right side of the inequality to\[\frac{\Varid{ecc}\ i\ j * (i - b) + 1 + 0 + 1 + \Varid{ecc}\ i\ j * (b + n - j)}{n}\ .\]

@@

With the right side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / n|, and then cancel the division by |n| on both sides. Next, to actually realize the split, we apply a specialized lemma stating that

>>>>

The right-hand inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
    was 
%%% could be 
fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.

The left inequality is the challenging one. To bring it closer to the index summation, we first write |l| on the left-hand side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).

@@

Again, the proof of (\ref{monoid_expec_IJ_lem}) 
is by induction on |qs|. But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
\begin{code}
  forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
    monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
\end{code}
In the base case, |li| is |nil|, and the left-hand side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
\begin{code}
monoid_expec (count (i, j)) (
    pi <- pick [0 ... n-1];
    let pivot := nth li pi in
    part <- partition_U pivot (remove li pi);
    lower <- qs_U (part Lt);
    upper <- qs_U (part Gt);
    ret (lower ++ pivot :: part Eq ++ upper)
  ) <= ecc i j.
\end{code}
Since |cmp_U| is deterministic, |partition_U| is as well. Furthermore, since
we know exactly what monadic effects |partition_U| has, we can split those
effects off and revert to simple effect-free |filter| passes. Finally, we rewrite using the following |monoid_expec| identity:
\begin{code}
  monoid_expec f (pick l >>= m) = avg (map (monoid_expec f fcmp m) l).
\end{code}
This way, the goal ends up in a form using less monadic indirection:
\begin{code}
avg (map (monoid_expec (count (i, j)) fcmp (fun pi =>
    let pivot := nth li pi in
    let rest := remove li pi in
    let flt := fun c => filter ((= c) fcmp lookup_cmp pivot) rest in
    ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
      lower <- qs_U (flt Lt);
      upper <- qs_U (flt Gt);
      ret (lower ++ pivot :: flt Eq ++ upper)
    ))) [0...n-1]) <= ecc i j.
\end{code}
Here, |map_fst| applies a function to a pair's first component.

We now distinguish between five different cases that can occur for the nondeterministically picked |pivot| (which, because we are in the |U| monad, is an index). It can either be less than |i|, equal to |i|, between |i| and |j|, equal to |j|, or greater than |j|. Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls). To represent this split, we first rewrite the right-hand side of the inequality to\[\frac{\Varid{ecc}\ i\ j * (i - b) + 1 + 0 + 1 + \Varid{ecc}\ i\ j * (b + n - j)}{n}\ .\]

@@

With the right-hand side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / n|, and then cancel the division by |n| on both sides. Next, to actually realize the split, we apply a specialized lemma stating that
